Loop-Aware Optimizations in PyPy's Tracing JIT

### 2012-08-06 talk/dls2012/paper.tex
use microtype
use footnote after full stop, protect space before citation

\usepackage{ifthen}
\usepackage{fancyvrb}
\usepackage{color}
\usepackage{ulem}
\usepackage{xspace}
\usepackage{epsfig}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage[utf8]{inputenc}
\usepackage{setspace}
\usepackage{relsize}

@@

The work described in this paper was done in the context of the PyPy
project\footnote{\texttt{http://pypy.org}}. PyPy is a framework for implementing
dynamic languages efficiently \cite{armin_rigo_pypys_2006}. When implementing a
language with PyPy, one writes an interpreter for the language in RPython
\cite{davide_ancona_rpython:_2007}. RPython (``Restricted Python``) is a subset
of Python chosen in such a way that it can be efficiently translated to a
C-based VM by performing type inference.

Many low-level aspects of the final VM are not contained within the interpreter
implementation but are inserted during translation to C. Examples for this are a
garbage collector and also a tracing JIT compiler \cite{bolz_tracing_2009}.

@@

The last section gave a motivating but unrealistically small example.
This section will define a slightly larger example that the rest of the paper
uses to demonstrate the effect of optimizations.
For this we are going to use a tiny interpreter for a dynamic language with
 a very small object
model, that just supports an integer and a float type (this example has been taken from a previous paper \cite{bolz_allocation_2011}). The objects support only
one operation, \lstinline{add}, which adds two objects (promoting ints to floats in a
mixed addition). The implementation of \lstinline{add} uses classical
double-dispatching.
%These classes could be part of the implementation of a very
%simple interpreter written in RPython.
The classes can be seen in
Figure~\ref{fig:objmodel} (written in RPython).

@@

If a pure operation appears more than once in the trace with the same input
arguments, it only needs be executed the first time and then the result
can be reused for all other appearances. PyPy's optimizers can also remove
repeated heap reads if the intermediate operations cannot have changed their
value\footnote{We perform a type-based alias analysis to know which
writes can affect which reads \cite{XXX}. In addition writes on newly allocated objects
can never change the value of old existing ones.}.

@@

PyPy's allocation removal optimization \cite{bolz_allocation_2011} makes it
possible to identify objects that are allocated within the loop but never
escape it. That is, no outside
object ever gets a reference to them. This
is performed by processing the operations in order and
optimistically removing every \lstinline{new} operation. Later on if
it is discovered that a reference to the object escapes the loop, the
\lstinline{new} operation is inserted at this point. All operations
(\lstinline{get}, \lstinline{set} and \lstinline{guard}) on the removed objects
are also removed and the optimizer needs to keep track of the value of all used
attributes of the object.

@@

In the general case, each allocation-removed object in the jump arguments is exploded into a
vector of variables containing the values of all registered
attributes\footnote{This is sometimes called \emph{scalar replacement}.}.
If some of the attributes are themselves references to
allocation-removed objects they are recursively exploded
to make the vector contain only concrete variables. Some care has
to be taken to always place the attributes in the same order when
performing this explosion. Notation becomes somewhat simpler if also every
concrete variable of the jump arguments is exploded into a vector containing
itself. For
every variable, $J_k$, of the original jump arguments, $J$, let
\begin{equation}
  \tilde J^{\left(k\right)} = \left\{
      \begin{array}{ll}
        \left(J_k\right)  & \text{if $J_k$ is concrete} \\
        H^{\left(k\right)} & \text{if $J_k$ is allocation-removed}
      \end{array}
  \right.
  ,
\end{equation}
where $H^{\left(k\right)}$ is a vector containing all concrete
attributes of $J_k$. The arguments of the optimized \lstinline{jump}
operation are constructed as the concatenation all the $\tilde J^{\left(k\right)}$ vectors,
\begin{equation}
  \hat J = \left( 
    \begin{array}{cccc}
      \tilde J^{\left(1\right)} & \tilde J^{\left(2\right)} & \cdots &
      \tilde J^{\left(|J|\right)} \\
    \end{array}
  \right)      
  .
\end{equation}
The arguments of the \lstinline{jump} operation of the peeled loop,
$K$, is constructed from $\hat J$ using the map $m$,
\begin{equation}
  \hat K = \left(m\left(\hat J_1\right), m\left(\hat J_1\right), 
                 \cdots, m\left(\hat J_{|\hat J|}\right)\right)
  .
\end{equation}
In the optimized trace $I$ is replaced by $\hat I$ and $K$ by $\hat
K$. The trace from Figure~\ref{fig:unopt-trace} will be optimized into

@@

We can observe that PyPy (even without loop peeling) is orders of magnitude
faster than either CPython or Psyco. This is due to the JIT compilation
advantages and optimizations we discussed in previous work
\cite{bolz_allocation_2011, bolz_runtime_2011}. The geometric mean of the
speedup of loop peeling is 70\%, which makes benchmark times
comparable with native-compiled C code. We attribute the performance gap to C code to
the relative immaturity of PyPy's JIT assembler backend as well as missing
optimizations, like instruction scheduling.

Other interesting interpreters that are helped greatly by this optimization are
for example our Prolog interpreter written in RPython
\cite{carl_friedrich_bolz_towards_2010}. Prolog programs often contain tight
loops that perform list processing. Furthermore we experimented with a Python library
for writing numerical kernels doing array manipulation. The exact extent is
out of scope for this paper.

@@

Mike Pall, the author of LuaJIT\footnote{\texttt{http://luajit.org/}} seems to
have developed the described technique independently. There are no papers about
LuaJIT but the author of it writes on a mailing list: ``The LOOP pass does
synthetic unrolling of the recorded IR, combining copy-substitution with
redundancy elimination to achieve code hoisting. The unrolled and
copy-substituted instructions are simply fed back into the compiler pipeline,
which allows reuse of all optimizations for redundancy elimination. Loop
recurrences are detected on-the-fly and a minimized set of PHIs is generated.''
\cite{pall_luajit_2009}

Both the Hotpath VM \cite{gal_hotpathvm:_2006} and SPUR
\cite{bebenita_spur:_2010} implements loop-invariant code motion
directly, by explicitly marking as loop-invariant all variables that stay the
same along all looping paths and then moving all pure computation that depends
only on these variables out of the loop. SPUR can also hoist loads out of the
loop if nothing in the loop can ever write to the memory location. It can also
move allocations out of the loop, but does not replace the object by its attributes.
This saves only the allocation, not the access to the object attributes.

The type specialization described by Gal \etal \cite{gal_trace-based_2009} can
be seen as doing a similar optimization (again by manually implementing it)
than the one described in Section~\ref{sub:allocation}: The effect of both is
that type checks are fully done before a loop is even entered.

>>>>

\usepackage{microtype}
\usepackage{ifthen}
\usepackage{fancyvrb}
\usepackage{color}
\usepackage{ulem}
\usepackage{xspace}
\usepackage{epsfig}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage[utf8]{inputenc}
\usepackage{setspace}
\usepackage{relsize}

@@

The work described in this paper was done in the context of the PyPy
project.\footnote{\texttt{http://pypy.org}} PyPy is a framework for implementing
dynamic languages efficiently~\cite{armin_rigo_pypys_2006}. When implementing a
language with PyPy, one writes an interpreter for the language in RPython~\cite{davide_ancona_rpython:_2007}.
RPython (``Restricted Python``) is a subset
of Python chosen in such a way that it can be efficiently translated to a
C-based VM by performing type inference.

Many low-level aspects of the final VM are not contained within the interpreter
implementation but are inserted during translation to C. Examples for this are a
garbage collector and also a tracing JIT compiler~\cite{bolz_tracing_2009}.

@@

The last section gave a motivating but unrealistically small example.
This section will define a slightly larger example that the rest of the paper
uses to demonstrate the effect of optimizations.
For this we are going to use a tiny interpreter for a dynamic language with
 a very small object
model, that just supports an integer and a float type (this example has been taken from a previous paper~\cite{bolz_allocation_2011}). The objects support only
one operation, \lstinline{add}, which adds two objects (promoting ints to floats in a
mixed addition). The implementation of \lstinline{add} uses classical
double-dispatching.
%These classes could be part of the implementation of a very
%simple interpreter written in RPython.
The classes can be seen in
Figure~\ref{fig:objmodel} (written in RPython).

@@

If a pure operation appears more than once in the trace with the same input
arguments, it only needs be executed the first time and then the result
can be reused for all other appearances. PyPy's optimizers can also remove
repeated heap reads if the intermediate operations cannot have changed their
value.\footnote{We perform a type-based alias analysis to know which
writes can affect which reads~\cite{XXX}. In addition writes on newly allocated objects
can never change the value of old existing ones.}

@@

PyPy's allocation removal optimization~\cite{bolz_allocation_2011} makes it
possible to identify objects that are allocated within the loop but never
escape it. That is, no outside
object ever gets a reference to them. This
is performed by processing the operations in order and
optimistically removing every \lstinline{new} operation. Later on if
it is discovered that a reference to the object escapes the loop, the
\lstinline{new} operation is inserted at this point. All operations
(\lstinline{get}, \lstinline{set} and \lstinline{guard}) on the removed objects
are also removed and the optimizer needs to keep track of the value of all used
attributes of the object.

@@

In the general case, each allocation-removed object in the jump arguments is exploded into a
vector of variables containing the values of all registered
attributes.\footnote{This is sometimes called \emph{scalar replacement}.}
If some of the attributes are themselves references to
allocation-removed objects they are recursively exploded
to make the vector contain only concrete variables. Some care has
to be taken to always place the attributes in the same order when
performing this explosion. Notation becomes somewhat simpler if also every
concrete variable of the jump arguments is exploded into a vector containing
itself. For
every variable, $J_k$, of the original jump arguments, $J$, let
\begin{equation}
  \tilde J^{\left(k\right)} = \left\{
      \begin{array}{ll}
        \left(J_k\right)  & \text{if $J_k$ is concrete} \\
        H^{\left(k\right)} & \text{if $J_k$ is allocation-removed}
      \end{array}
  \right.
  ,
\end{equation}
where $H^{\left(k\right)}$ is a vector containing all concrete
attributes of $J_k$. The arguments of the optimized \lstinline{jump}
operation are constructed as the concatenation all the $\tilde J^{\left(k\right)}$ vectors,
\begin{equation}
  \hat J = \left( 
    \begin{array}{cccc}
      \tilde J^{\left(1\right)} & \tilde J^{\left(2\right)} & \cdots &
      \tilde J^{\left(|J|\right)} \\
    \end{array}
  \right)      
  .
\end{equation}
The arguments of the \lstinline{jump} operation of the peeled loop,
$K$, is constructed from $\hat J$ using the map $m$,
\begin{equation}
  \hat K = \left(m\left(\hat J_1\right), m\left(\hat J_1\right), 
                 \cdots, m\left(\hat J_{|\hat J|}\right)\right)
  .
\end{equation}
In the optimized trace $I$ is replaced by $\hat I$ and $K$ by $\hat
K$. The trace from Figure~\ref{fig:unopt-trace} will be optimized into

@@

We can observe that PyPy (even without loop peeling) is orders of magnitude
faster than either CPython or Psyco. This is due to the JIT compilation
advantages and optimizations we discussed in previous
work~\cite{bolz_allocation_2011, bolz_runtime_2011}. The geometric mean of the
speedup of loop peeling is 70\%, which makes benchmark times
comparable with native-compiled C code. We attribute the performance gap to C code to
the relative immaturity of PyPy's JIT assembler backend as well as missing
optimizations, like instruction scheduling.

Other interesting interpreters that are helped greatly by this optimization are
for example our Prolog interpreter written in
RPython~\cite{carl_friedrich_bolz_towards_2010}. Prolog programs often contain
tight
loops that perform list processing. Furthermore we experimented with a Python library
for writing numerical kernels doing array manipulation. The exact extent is
out of scope for this paper.

@@

Mike Pall, the author of LuaJIT\footnote{\texttt{http://luajit.org/}} seems to
have developed the described technique independently. There are no papers about
LuaJIT but the author of it writes on a mailing list: ``The LOOP pass does
synthetic unrolling of the recorded IR, combining copy-substitution with
redundancy elimination to achieve code hoisting. The unrolled and
copy-substituted instructions are simply fed back into the compiler pipeline,
which allows reuse of all optimizations for redundancy elimination. Loop
recurrences are detected on-the-fly and a minimized set of PHIs is
generated.''~\cite{pall_luajit_2009}

Both the Hotpath VM~\cite{gal_hotpathvm:_2006} and
SPUR~\cite{bebenita_spur:_2010} implements loop-invariant code motion
directly, by explicitly marking as loop-invariant all variables that stay the
same along all looping paths and then moving all pure computation that depends
only on these variables out of the loop. SPUR can also hoist loads out of the
loop if nothing in the loop can ever write to the memory location. It can also
move allocations out of the loop, but does not replace the object by its attributes.
This saves only the allocation, not the access to the object attributes.

The type specialization described by Gal \etal~\cite{gal_trace-based_2009} can
be seen as doing a similar optimization (again by manually implementing it)
than the one described in Section~\ref{sub:allocation}: The effect of both is
that type checks are fully done before a loop is even entered.
